---
title: "Bacterial richness in public tap water and sinks of San Francisco's public Pit Stop bathrooms"
author: "Andrae Ladores"
date: "April 4, 2022"
bibliography: references.bib
csl: ecology.csl
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Throughout the world and including the United States, there are numerous individuals that are deprived of safe and reliable tap water. Cities like Newark, New Jersey and Flint, Michigan are only some examples of the water crisis in the United States [@doi:10.2105/AJPH.2021.306561]. In California, there is a small percentage of public tap water systems that have been found to be in violation at least one point in their time since 2012 [@reese2018does]. Further studies have revealed that water system violations in California transpire in low-income communities, where there are limited resources which makes it difficult to meet the standards, driving the inequitable access to safe tap water [@balazs2012environmental; @doi:10.2105/AJPH.2013.301664]. 
Besides access to safe public water, public restrooms are a necessity as well. They provide service to a mobile population and assists the individuals with greater needs (children, elderly, the homeless, etc.) [@STANWELLSMITH2010613]. In addition, public restrooms in high-income locations are known to display cleaner facilities and conditions compared to the restrooms in low and medium-income areas [@suen2019public].
In this study, I focused on the bacterial richness in public tap water of San Francisco's public restrooms (Pit Stop bathrooms) and how they vary regarding population density and the average neighborhood income. Additionally, I tested the pH levels between Outer Richmond's Lake Stow and Lakeshore's Lake Merced to investigate the pH levels between the lakes with differing green space. The study revolves around the use of field collected data across several districts of San Francisco and the use of pre-existing datasets relating to the study of different urban systems. The field collected data consisted of tap water samples of Pit Stop bathrooms along with swab samples of several locations on the sink in the same bathrooms. The publicly-available datasets consisted of information pertaining to San Francisco's average income values along with San Francisco's census data.  

# Methods

Add text here.

## Site Locations and Descriptions

Add text here.

## Field Sampling Design

Add text here.

## Data Analysis and Statistics

Add text here.

# Results

## Subsections are ok in the results section too

Add a number of code chunks in the Results section. These should read in, subset and plot the data as needed (no need to save any figures to pdf, since they will be put into the rendered document when you click 'knit'), and, for any hypotheses that you want to test, an appropriate statistical test. 

```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
```

# Discussion

# Sources Cited

