---
title: "Bacterial richness in public tap water and sinks of San Francisco's public Pit Stop bathrooms"
author: "Andrae Ladores"
date: "April 4, 2022"
bibliography: references.bib
csl: ecology.csl
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Throughout the world and including the United States, there are numerous individuals that are deprived of safe and reliable tap water. Cities like Newark, New Jersey and Flint, Michigan are only some examples of the water crisis in the United States [@doi:10.2105/AJPH.2021.306561]. In California, there is a small percentage of public tap water systems that have been found to be in violation at least one point in their time since 2012 [@reese2018does]. Further studies have revealed that water system violations in California transpire in low-income communities, where there are limited resources which makes it difficult to meet the standards, driving the inequitable access to safe tap water [@balazs2012environmental; @doi:10.2105/AJPH.2013.301664].

# Methods

Add text here.

## Site Locations and Descriptions

Add text here.

## Field Sampling Design

Add text here.

## Data Analysis and Statistics

Add text here.

# Results

## Subsections are ok in the results section too

Add a number of code chunks in the Results section. These should read in, subset and plot the data as needed (no need to save any figures to pdf, since they will be put into the rendered document when you click 'knit'), and, for any hypotheses that you want to test, an appropriate statistical test. 

```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
```

# Discussion

# Sources Cited

